---
title: "Assignment 5: Data Visualization"
author: "Jordan Mullens"
date: "Spring 2023"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
chunk_output_type: inline
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1.  Rename this file `Jordan_Mullens_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Upload the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 Loading packages
library(ggplot2)
library(lubridate)
library(tidyverse)
library(here)
here()

PeterPaul.chem.nutrients <- 
  read.csv(here("Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"),
           stringsAsFactors = T)
PeterPaul.Niwot.Litter <- 
  read.csv(here("Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv"),
           stringsAsFactors = T)

#2 
PeterPaul.chem.nutrients$sampledate <- ymd(PeterPaul.chem.nutrients$sampledate)
PeterPaul.Niwot.Litter$collectDate <- ymd(PeterPaul.Niwot.Litter$collectDate)
class(PeterPaul.chem.nutrients$sampledate)

```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3
library(ggthemes)
my_theme <- theme_base() + 
  theme(
    line = element_line(
      color='black',
      linewidth =2
    ),
    legend.background = element_rect(
      color='grey',
      fill = 'green'
    ),
    legend.title = element_text(
      color='blue'
    ))


```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 
peterpaul.Phos.plot1 <- 
  ggplot(PeterPaul.chem.nutrients, aes(
         x = po4,
         y = tp_ug,
         color = lakename)) +
  geom_point() +
  geom_smooth(method=lm, se=FALSE, color= "black") +
   xlim(0, 40) +
  ylim(0, 110) +
  facet_wrap(vars(lakename)) +
  labs(
    title = "Peter Lake and Paul Lake Phosphorous vs Phosphate",
    y= "Phosphorous Concentration",
    x= "Phosphate Concentration",
    color= "year") 

print(peterpaul.Phos.plot1)



```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tip: R has a build in variable called `month.abb` that returns a list of months;see <https://r-lang.com/month-abb-in-r-with-example>

```{r Create boxplots}
#5 
#Temperature Box Plot
peterpaul.Temp.box.plot <- 
  ggplot(PeterPaul.chem.nutrients, aes(
         x = factor(month),
         y = temperature_C,
         color = lakename)) +
  geom_boxplot() +
  labs(
    title = "Temperature vs Month",
    y= "Temperature in Degrees C",
    x= "Month") 

print(peterpaul.Temp.box.plot)


#TP Box Plot
peterpaul.TP.box.plot <- 
  ggplot(PeterPaul.chem.nutrients, aes(
         x = factor(month),
         y = tp_ug,
         color = lakename)) +
  geom_boxplot() +
  labs(
    title = "TP vs Month",
    y= "TP",
    x= "Month") 

print(peterpaul.TP.box.plot)


#TN Box Plot
peterpaul.TN.box.plot <- 
  ggplot(PeterPaul.chem.nutrients, aes(
         x = factor(month),
         y = tn_ug,
         color = lakename)) +
  geom_boxplot() +
  labs(
    title = "TN vs Month",
    y= "TN",
    x= "Month") 

print(peterpaul.TN.box.plot)


#Creating a cowplot
library(cowplot)

legend_cow <- get_legend(
  peterpaul.TN.box.plot + 
  guides(color = guide_legend(nrow = 1)) +
  theme(legend.position = "right"))



peter.paul.cowplot <- plot_grid(
  peterpaul.Temp.box.plot + theme(legend.position="none"),
  peterpaul.TN.box.plot + theme(legend.position="none"),
  peterpaul.TP.box.plot + theme(legend.position="none"),
  hjust = -1,
  nrow = 1,
  legend_cow
  )
print(peter.paul.cowplot)



```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: Temperature peaks in the summer months for both lakes. TP peaks in the summer for Peter Lake and is at a minimum for Paul lake. TN appears to be relatively consistent throughout the year.

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6
Niwot.Ridge.plot1.litter <- PeterPaul.Niwot.Litter %>%
  filter(functionalGroup == "Needles") %>% 
  ggplot( 
    aes(
         x = collectDate,
         y = dryMass,
         color = nlcdClass)) +
  geom_point() +
  labs(
    title = "Niwot Ridge Dry Mass vs Collection Date",
    y= "Dry Mass",
    x= "Collection Date",
    color= "NLCD Class")

print(Niwot.Ridge.plot1.litter)


#7
Niwot.Ridge.plot1.litter <- PeterPaul.Niwot.Litter %>%
  filter(functionalGroup == "Needles") %>% 
  ggplot( 
    aes(
         x = collectDate,
         y = dryMass,
         color = nlcdClass)) +
  geom_point() +
  facet_wrap(vars(nlcdClass)) +
  labs(
    title = "Niwot Ridge Dry Mass vs Collection Date",
    y= "Dry Mass",
    x= "Collection Date",
    color= "NLCD Class")

print(Niwot.Ridge.plot1.litter)



```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: The second one is more effective because it's easier to see variations within each NLCD class. 



